Informative Oceanographic Sampling using Excursion Probabilities for Multivariate Random Fields

### 2019-09-15 main_new.tex
updates

\subsection{Illustration}

@@

Table \ref{tab:sim_rhoab} shows the EPs and the associated Bernoulli
variance (second row) for the examples indicated in
Fig. \ref{illus_bivarDens}. The EPs increase with the correlation as
there is a stronger tendency of having jointly small temperature and
salinity. The Bernoulli variance is similarly largest for high
correlation. EPs and Bernoulli variances are the same for standard
deviation $1$ or $2$, and this means that high variability in the
temperature and salinity variables is not captured in the $p(1-p)$
expression.

\begin{table}[!h] \centering \caption{EP and Bernoulli variance for
    different correlations and variances (top rows), and expected
    Bernoulli variances for both data \kc{Do you mean an update of the
      data?} and only for temperature (bottom
    rows).}
  \begin{tabular}{c|ccc|ccc}
 &\multicolumn{3}{c}{$\sigma_b=\sigma_a=1$} & \multicolumn{3}{c}{$\sigma_b=\sigma_a=2$} \\
\hline
Correlation $\gamma$ & 0.2 & 0.6 & 0.8 & 0.2 & 0.6 & 0.8 \\
\hline
$p$ & 0.28 & 0.35 & 0.40 & 0.28 & 0.35 & 0.40 \\ 
$p(1-p)$ & 0.20 & 0.23 & 0.24 & 0.20 & 0.23 & 0.24 \\ 
$E_{(y_a,y_b)}(p (1-p))$ & 0.092 & 0.089 & 0.085 & 0.052 & 0.051 & 0.049 \\ 
$E_{y_a}(p (1-p))$ & 0.151 & 0.138 & 0.123 & 0.137 & 0.114 & 0.093 \\ 
\hline
\end{tabular}
\label{tab:sim_rhoab}
\end{table}

Table \ref{tab:sim_rhoab} (bottom two rows) shows results of expected Bernoulli variance calculations. This is presented for a design gathering both data types $(y_a,y_b)$, and for a design with temperature measurements $y_a$ alone. Having both data; $(y_a,y_b)^t=(\xi_a,\xi_b)^t+N(0,0.5^2\bI)$, while $y_a=\xi_a+N(0,0.5^2)$ when only temperature is measured.
The results in Table \ref{tab:sim_rhoab} shows that 
the expected Bernoulli variance gets lower with larger standard deviations $\sigma_a$ and $\sigma_b$ (right columns). The reduction of Bernoulli variance is largest for the cases with high correlation $\gamma$. Albeit smaller, there is also uncertainty reduction when only temperature is measured (bottom row), especially when the temperature and salinity are very dependent. When correlation is low ($\gamma=0.2$) there is hardly any information about salinity in the temperature data, and hence less uncertainty reduction. In the application with fresh cold water from the river, the temperature and salinity variables will not only be interdependent, but they will be dependent in the spatial dimension. This dependence will also impact design criteria when we evaluate the information measure by integrating over several locations $\bx$.

@@

Recall that the AUV can adapt its survey plan based on what it is measuring, in accordance with the sense, plan, act loop (Fig. \ref{fig:sense-plan-act}). Here we present adaptive strategies for sampling.

@@

An adaptive AUV survey is split in many rounds (stages), and at each round, the survey design is found by considering the largest reduction in uncertainty of the ES. In our setting the selection is made on the graph defined by the nearest grid nodes in the domain $\mathcal{M}_g$.

The optimal sequential design does not only consider the best current AUV grid node, but also what data gathered at this node would lead to for future sampling at the next stages. 
Let $d^{j,s}$ denote design number $j$ at stage $s$ of sequential surveying. If this design is selected, data $\by^{j,s}$ will be gathered. The optimal path selection situation is then depicted in Figure \ref{fig:PathSelOpt}, 
\begin{figure}[h!]
\centering
\includegraphics[width=0.85\textwidth]{Figures/sequent_select.pdf}
\caption{Optimal sequential path selection.}\label{fig:PathSelOpt}
\end{figure}
where design choices are indicated by squares while data realizations are indicated by circles. 
%terms this optimal design is defined as
%\begin{equation}\label{opt_crit}
%    \bd^* = \mbox{argmin}_{d^{j,1}} \left\{ \int_{\by^{j,1}} \mbox{argmin}_{d^{j,2}} \left\{ \int_{\by^{j,2}} \ldots \pi(\by^{j,2}|\by^{j,1}) d\by^{j,2} \right\} \pi(\by^{j,1}) d \by^{j,1} \right\},
%\end{equation}
%where $\ldots$ represents the expected variance reduction in the ES under the continued optimal design, which will depend on the data at earlier stages. 
The mathematical expression for the optimal design then involves a series of intermixed maximizations over designs and integrals over data.
In practice the optimal solution is not tractable because of the enormous growth over stages, see e.g. \cite{sucar2015probabilistic} and \cite{powell2016perspectives}.

Instead, we next outline heuristic strategies. In terms of notation, let $\mathcal{Y}^{s-1} =\{d^{j,r}, \by^{j,r} ; r=1,\ldots,s-1 \}$ denote the data gathered until stage $s-1$, using a selected design $d^{j,r}$, $r=1,\ldots,s-1$.

The derived closed form solutions for the expected IBV are still important building blocks when constructing the sampling designs as they will be used to score the different adaptive design. Efficient calculation is important for using adaptive survey designs with robotic platforms.

\subsection{Naive path planning}
\label{naive}

The simplest heuristic for adaptive sampling is to choose the next sampling location based on current EPs. The variance is largest for EPs equal to $0.5$, so one selects the design location with EP closest to $0.5$.

At stage $s$, based on the currently available data $\mathcal{Y}^{s-1}$, we fit an updated Gaussian model from Eq. (\ref{gp_upd}), with mean $\bm_{s-1}=E(\bxi|\mathcal{Y}^{s-1})$ and covariance matrix $\bS_{s-1}=\mbox{Var}(\bxi|\mathcal{Y}^{s-1})$. 
The next round of measurements $\by^{j,s}$, can be gathered with design $d^{j,s}$, $j=1,\ldots,J$, where $J$ indicates all possible AUV directions or paths at the current stage. The {\it{naive}} strategy selects the design according to
\begin{eqnarray}\label{critNaive}
    d^{*,s} &=& \mbox{argmin}_{j \in \{1,\ldots,J\}} |p(\bx_{d^{j,s}};\mathcal{Y}^{s-1})-0.5|, \\
    p(\bx;\mathcal{Y}^{s-1}) &=& P(\xi_a(\bx) \leq t_a, \xi_b(\bx) \leq t_b | \mathcal{Y}^{s-1}). \nonumber
\end{eqnarray}
This strategy does not account for the uncertainty in the temperature or salinity variables, only if one design (node) has EPs closer to $0.5$ (see Table \ref{tab:sim_rhoab}, line two). Neither does it account for spatial correlation. This strategy lacks memory of where it has been and where the uncertainty has been reduced. For this reason it can easily get stuck in local regions.

@@

Based on the currently available data $\mathcal{Y}^{s-1}$, we again fit an updated Gaussian model, represented on the grid locations covering the spatial domain. 
The next round of measurements $\by^{j,s}$, can be gathered with design $d^{j,s}$, $j=1,\ldots,J$, indicating all possible AUV directions / paths at the current stage. The selected design is then
\begin{eqnarray}\label{critSEQ}
    d^{*,s} &=& \mbox{argmin}_{j \in \{1,\ldots,J\}} \left\{ V_{m,\mbox{upd,j}} \right\},  \\
V_{m,\mbox{upd}} & \approx & \sum_{\bx \in \mathcal{M}_g} E_{\by^{j,s}|\mathcal{Y}^{s-1}} \left\{ p(\bx;\mathcal{Y}^{j,s})\left( 1-p(\bx;\mathcal{Y}^{j,s})\right) \right\} \Delta, \nonumber \\
    p(\bx;\mathcal{Y}^{j,s}) &=& P(\xi_a(\bx) \leq t_a, \xi_b(\bx) \leq t_b |\by^{j,s},\mathcal{Y}^{s-1}). \nonumber
\end{eqnarray}

Note that this strategy gives a sequential conditional version of the formula in Eq. (\ref{sur}). Now $\mathcal{Y}^{s-1}$ is available, and the expectation is with respect to the conditional density $\pi(\by^{j,s}|\mathcal{Y}^{s-1})$. A similar closed form calculation for expected IBV is hence applicable in Eq. (\ref{critSEQ}), using the updated GP model from step $s-1$. 
Once the data are collected for the best design, the GP model is updated again. The mean $\bm_{s}$ and covariance matrix $\bS_{s}$ are used to compute the next design at stage $s+1$, and so on.

Even though this myopic strategy is non-anticipative, it still gives a reasonable approach for creating designs in many applications. Moreover, it is easily implemented on-board an AUV, using the efficient approach for data updating of the GP model and the calculation of the closed form expected IBV expressions for each next survey line.

\subsection{Look-ahead path planning}
\label{LA}

We now extend the myopic strategy to a look-ahead strategy which is optimal when one can gather only two more rounds of measurements. In addition to the next round of measurements $\by^{j,s}$, this look-ahead strategy hence anticipates the subsequent design $j_2$ with data $\by^{j_2,s+1}$, when choosing the current design $d^{j,s}$. 
The selected design is
\begin{eqnarray}\label{critLA}
    d^{*,s} &=& \mbox{argmin}_{j \in \{1,\ldots,J\}} \left\{ U_{la,\mbox{upd,j}} \right\},  \\
    U_{la,\mbox{upd},j} & = &  E_{\by^{j,s}|\mathcal{Y}^{s-1}} \left\{ \mbox{argmin}_{j_2 \in \{1,\ldots,{J}_2\}} \left[ V_{la,\mbox{upd},j_2} \right] \right\}, \nonumber \\
V_{la,\mbox{upd},j_2} & \approx & \sum_{\bx \in \mathcal{M}_g} E_{\by^{j_2,s+1}|\mathcal{Y}^{j,s}} \left\{ p(\bx;\mathcal{Y}^{j,s+1})\left( 1-p(\bx;\mathcal{Y}^{j,s+1})\right) \right\} \Delta, \nonumber \\
    p(\bx;\mathcal{Y}^{j,s+1}) &=& P(\xi_a(\bx) \leq t_a, \xi_b(\bx) \leq t_b |\by^{j_2,s+1},\mathcal{Y}^{j,s}). \nonumber
\end{eqnarray}
Here, $\mathcal{Y}^{j,s}=\{\by^{j,s},\mathcal{Y}^{s-1}\}$ and $\mathcal{Y}^{j,s+1}=\{\by^{j_2,s+1},\mathcal{Y}^{j,s}\}$ represent the sets of data variables, and the expectations are with respect to the conditional densities $\pi(\by^{j,s}|\mathcal{Y}^{s-1})$ and $\pi(\by^{j_2,s+1}|\by^{j,s},\mathcal{Y}^{s-1})$.

@@

Even though the strategy looks at two stages, it is only used to find the current best design. When data are collected, the GP model is updated, and the mean $\bm_{s}$ and covariance matrix $\bS_{s}$ are used to compute the next design at stage $s+1$, now anticipating what stage $s+2$ could offer, and so on.

This look-ahead approach is much more computer-demanding than the myopic strategy, and for practical implementation we prune paths in the evaluation of Eq. (\ref{critLA}). This means that we do not compute all possible branches of the first two stages, as they are indicated in Figure \ref{fig:PathSelOpt}. Instead, we use the myopic strategy to rank the three best designs on the first stage alone, and for each of these we go through with the look-ahead calculation.

\section{Simulation study}\label{sec:simulations}

Results from tests performed to study the properties of different static and sequential survey designs, in a realistic simulated case, is shown in this section. The context is mapping a freshwater plume defined by a temperature and salinity field using an AUV.

@@

We use a bivariate GP model for temperature and salinity, where we specify the prior mean 
\begin{equation}\label{m}
    \bmu(\bx)=E 
    \begin{bmatrix}
    \xi_a(\bx) \\
    \xi_b(\bx) 
    \end{bmatrix}=\begin{bmatrix} \mu_{a}(\bx)\\
\mu_{b}(\bx)
\end{bmatrix} 
= \begin{bmatrix} \beta_{a,0} + \beta_{a,1} x_{\mbox{East}} \\
\beta_{a,1} + \beta_{b,1} x_{\mbox{East}}
\end{bmatrix}.
\end{equation}
In this simulation setup, which is motivated by the real case, we expect the eastern part of the domain to be cold freshwater. This situation mimics that of a river mouth entering from the south, and the water masses are pulled to the east. The actual trends are in practice set from preliminary information or runs of oceanographic models solving the differential equations for fluid flow. In this example the mean is specified by $\beta_{a,1}=0.065$ and $\beta_{b,1}=0.1$. Further, the intercepts of the temperature and salinity variables are set to $\beta_{a,0}=5.8$ and $\beta_{b,1}=29.0$.

@@

The covariance is assumed to be stationary and separable for the two variables. 
The $2 \times 2$ pointwise covariance matrix is set to
\begin{equation}\label{v0}
\bSigma(\bx)=\mbox{Var} 
\begin{bmatrix}
    \xi_a(\bx) \\
    \xi_b(\bx) 
    \end{bmatrix}=
\begin{bmatrix}
0.25^2 & 0.6 \cdot 0.25^2 \\
0.6 \cdot 0.25^2 & 0.25^2
\end{bmatrix}.
\end{equation}
The spatial correlation is of a Matern type giving
\begin{equation}\label{v}
\mbox{Corr} 
\left(
\begin{bmatrix}
    \xi_a(\bx) \\
    \xi_b(\bx) 
    \end{bmatrix},
    \begin{bmatrix}
    \xi_a(\bx') \\
    \xi_b(\bx') 
    \end{bmatrix}
    \right)
    = \begin{bmatrix}
1 & 0.6  \\
0.6  & 1
\end{bmatrix}(1+\phi |\bx-\bx'|)\exp (-\phi |\bx-\bx'|),
\end{equation}
and $\phi=0.3$ indicates an effective correlation range of about $1200$ m. 
%Figure \ref{fig:stat_design} shows the contour lines for the EP for the reference. We notice the trend of increasing salinity and temperature to the west.  
We study sensitivity to the parameter specification in the results below.
One realization of the random fields representing salt and temperature is shown in Figure \ref{fig:true_temp} and \ref{fig:true_sal}. The true ES from this prior realization is plotted in Figure \ref{fig:ESet}.

\begin{figure}[!h]
  \centering
  \subfigure[Simulated temperature.]{\includegraphics[height = 0.40\textwidth]{Figures/sim/true_temp.pdf}\label{fig:true_temp}}
  \hfill
  \subfigure[Simulated salinity.]{\includegraphics[height = 0.40\textwidth]{Figures/sim/true_sal.pdf}\label{fig:true_sal}}
  \hfill
  \subfigure[Excursion set given the realization of temperature and salinity.]{\includegraphics[height = 0.40\textwidth]{Figures/sim/es_ts_true.pdf}\label{fig:ESet}}
  \hfill
  \subfigure[Excursion probabilities given collected data along the strategy ``static\_north".]{\includegraphics[height = 0.40\textwidth]{Figures/sim/es_ts_posterior.pdf}\label{fig:Eprob}}
  \caption{\ref{fig:true_temp} and \ref{fig:true_sal} shows
    realisations of the simulated temperature and salinity fields, as
    well as the associated excursion set \ref{fig:ESet}.
    \ref{fig:Eprob} show the estimated excursion probabilities after
    performing data collection along the N-S survey line.} 
\label{fig:realisations}
\end{figure}

@@

We compare three different static designs denoted \texttt{static\_north}, \texttt{static\_east and} \texttt{static\_zigzag} with the three sequential approaches \texttt{naive}, \texttt{myopic}, and \texttt{look-ahead}. The static sampling paths are pre-determined and cannot be changed on the basis of data, representing the archetypical pre-planned observation strategies used in current observational practices. For a fixed survey length, the expected variance after static surveying is analytically available (equal coverage each time), but here we focus on the properties from the sequential designs which are evaluated using Monte Carlo sampling over several replicates of realizations from the prior, and conducting simulated sequential surveying for each one; making the result dependent on the observed data. For comparison, the same setup is used for the static designs. Figure \ref{fig:Eprob} shows the conditional EP, given data gathered along the north-south survey line for the realization indicated in these displays. In the Monte Carlo replicates, such results are repeatedly computed to approximate the expected variance reduction in the ES. Along with this criterion we also compare predictive performance measured by root mean square error (RMSE) in the temperature and salinity fields and variance reduction in these two field variables. It is important to note that the objective function used by the agents is focused on reducing the expected variance in the ES, but we nevertheless hope to achieve good predictive performance for criteria such as RMSE as well. Another non-statistical criteria that is important for practical purposes is the computing time of the strategy, as this will impact the performance of the approach in the field for an embedded system.

The sequential sampling agents start at the center East-West coordinate at the southern end of the domain (node 53), see Fig. \ref{fig:wp_graph}.

\begin{figure}[h]
\centering
\includegraphics[width=0.50\textwidth]{Figures/sim/wp_graph_paper.pdf}
\caption{The equilateral waypoint graph used to discretize the route choices over the $31\times31$ grid.}
\label{fig:wp_graph}
\end{figure}

Maintaining the AUV example, the AUV will move along edges in the Fig. \ref{fig:wp_graph} waypoint graph. While doing so, the AUV will gather measurements. These are assimilated in the GP model, before an evaluation of the next node to sample will be conducted at the end of the edge. The paths of will differ for the various sequential designs due to the individual design criteria and the simulated variability in the environment among the replicates.

@@

\section{Case study - Mapping a river plume}\label{sec:case_study}

>>>>

\subsection{An Example}

@@

Table \ref{tab:sim_rhoab} shows the EPs and the associated Bernoulli
variance (second row) for the examples indicated in
Fig. \ref{illus_bivarDens}. The EPs increase with the correlation as
there is a stronger tendency of having jointly small temperature and
salinity. The Bernoulli variance is similarly largest for high
correlation. EPs and Bernoulli variances are the same for standard
deviation $1$ or $2$, and this implies that high variability in
temperature and salinity is not captured in the $p(1-p)$ expression.

\begin{table}[!h] \centering \caption{EP and Bernoulli variance for
    different correlations and variances (top rows), and expected
    Bernoulli variances for both data \kc{Do you mean an update of the
      data? What is 'both' in this context? T and S? I believe its the
    latter, and so should be explicit.} and only for
    temperature (bottom rows).}
  \begin{tabular}{c|ccc|ccc}
 &\multicolumn{3}{c}{$\sigma_b=\sigma_a=1$} & \multicolumn{3}{c}{$\sigma_b=\sigma_a=2$} \\
\hline
Correlation $\gamma$ & 0.2 & 0.6 & 0.8 & 0.2 & 0.6 & 0.8 \\
\hline
$p$ & 0.28 & 0.35 & 0.40 & 0.28 & 0.35 & 0.40 \\ 
$p(1-p)$ & 0.20 & 0.23 & 0.24 & 0.20 & 0.23 & 0.24 \\ 
$E_{(y_a,y_b)}(p (1-p))$ & 0.092 & 0.089 & 0.085 & 0.052 & 0.051 & 0.049 \\ 
$E_{y_a}(p (1-p))$ & 0.151 & 0.138 & 0.123 & 0.137 & 0.114 & 0.093 \\ 
\hline
\end{tabular}
\label{tab:sim_rhoab}
\end{table}

Table \ref{tab:sim_rhoab} (bottom two rows) shows results of expected
Bernoulli variance calculations. This is presented for a design
gathering both data types $(y_a,y_b)$, and for a design with
temperature measurements $y_a$ alone. Having both data;
$(y_a,y_b)^t=(\xi_a,\xi_b)^t+N(0,0.5^2\bI)$, while
$y_a=\xi_a+N(0,0.5^2)$ when only temperature is measured.  The results
in Table \ref{tab:sim_rhoab} show that the expected Bernoulli variance
gets lower with larger standard deviations $\sigma_a$ and $\sigma_b$
(right columns) \kc{With just 2 data points isn't this a too strong
  statment to make?}. The reduction of Bernoulli variance is largest
for the cases with high correlation $\gamma$. Albeit smaller, there is
also uncertainty reduction when only temperature is measured (bottom
row), especially when temperature and salinity are very
dependent. When correlation is low ($\gamma=0.2$) there is little
information about salinity in the temperature data, and hence less
uncertainty reduction. In an application with fresh cold water from
river source, the temperature and salinity variables will not only be
interdependent, but will show dependence in the spatial
dimension. This in turn will impact the design criteria when we
evaluate the information measure by integrating over several locations
$\bx$.

@@

Recall AUV's can adapt survey plans based on what is being measured,
in accordance with the sense, plan, act loop
(Fig. \ref{fig:sense-plan-act}). Here we present adaptive strategies
for sampling.

@@

An adaptive AUV survey is split in many rounds (stages), and at each
round, the survey design is found by using a myopic approach of
considering the largest reduction in uncertainty of the ES. In our
setting the selection is made on the graph defined by the nearest grid
nodes in the domain $\mathcal{M}_g$.

The optimal sequential design not only considers the best current AUV
grid node, but also what data gathered at this node would lead to for
future sampling at successive stages.  Let $d^{j,s}$ denote design
number $j$ at stage $s$ of a sequential survey. If this design is
selected, data $\by^{j,s}$ will be gathered. The optimal path
selection situation is then depicted in Figure \ref{fig:PathSelOpt},
\begin{figure}[h!]
\centering
\includegraphics[width=0.85\textwidth]{Figures/sequent_select.pdf}
\caption{Optimal sequential path selection.}\label{fig:PathSelOpt}
\end{figure}
where design choices are indicated by squares while data realizations are indicated by circles. 
%terms this optimal design is defined as
%\begin{equation}\label{opt_crit}
%    \bd^* = \mbox{argmin}_{d^{j,1}} \left\{ \int_{\by^{j,1}} \mbox{argmin}_{d^{j,2}} \left\{ \int_{\by^{j,2}} \ldots \pi(\by^{j,2}|\by^{j,1}) d\by^{j,2} \right\} \pi(\by^{j,1}) d \by^{j,1} \right\},
%\end{equation}
%where $\ldots$ represents the expected variance reduction in the ES under the continued optimal design, which will depend on the data at earlier stages. 
The mathematical expression for the optimal design then involves a
series of intermixed maximizations over designs and integrals over
data.  In practice the optimal solution is not tractable because of
the enormous growth over stages, see
e.g. \cite{sucar2015probabilistic} and \cite{powell2016perspectives}.
Instead, we outline heuristic strategies. In terms of notation, let
$\mathcal{Y}^{s-1} =\{d^{j,r}, \by^{j,r} ; r=1,\ldots,s-1 \}$ denote
the data gathered until stage $s-1$, using a selected design
$d^{j,r}$, $r=1,\ldots,s-1$.

The derived closed form solutions for the expected IBV are still
important building blocks when constructing the sampling designs as
they will be used to score the different adaptive designs. % Efficient
% calculation is important for using adaptive survey designs with
% robotic platforms.

\subsection{A Naive Sampling Strategy}
\label{naive}

The simplest heuristic for adaptive sampling is to choose the next
sampling location based on current EPs. The largest variance in
sampling locations is and indication of higher uncertainty; therefore
one selects the design location with an EP closest to $0.5$.

At stage $s$, based on the currently available data
$\mathcal{Y}^{s-1}$, we fit an updated Gaussian model from
Eq. (\ref{gp_upd}), with mean $\bm_{s-1}=E(\bxi|\mathcal{Y}^{s-1})$
and covariance matrix $\bS_{s-1}=\mbox{Var}(\bxi|\mathcal{Y}^{s-1})$.
The next round of measurements $\by^{j,s}$, can be gathered with
design $d^{j,s}$, $j=1,\ldots,J$, where $J$ indicates all possible AUV
trajectories from the current stage. The {\it{naive}} strategy then,
selects the design according to:
\begin{eqnarray}\label{critNaive}
    d^{*,s} &=& \mbox{argmin}_{j \in \{1,\ldots,J\}} |p(\bx_{d^{j,s}};\mathcal{Y}^{s-1})-0.5|, \\
    p(\bx;\mathcal{Y}^{s-1}) &=& P(\xi_a(\bx) \leq t_a, \xi_b(\bx) \leq t_b | \mathcal{Y}^{s-1}). \nonumber
\end{eqnarray}
This strategy does not account for the uncertainty in the temperature
or salinity variables, only if one design (node) has EPs closer to
$0.5$ (see Table \ref{tab:sim_rhoab}, line two). Neither does it
account for spatial correlation. This strategy lacks memory of where
it has been and where the uncertainty has been reduced and therefore
susceptible to local minimas.

@@

Based on the currently available data $\mathcal{Y}^{s-1}$, we again
fit an updated Gaussian model, represented on the grid locations
covering the spatial domain.  The next round of measurements
$\by^{j,s}$, can be gathered with design $d^{j,s}$, $j=1,\ldots,J$,
indicating all possible AUV trajectories at the current stage. The
selected design is then:
\begin{eqnarray}\label{critSEQ}
    d^{*,s} &=& \mbox{argmin}_{j \in \{1,\ldots,J\}} \left\{ V_{m,\mbox{upd,j}} \right\},  \\
V_{m,\mbox{upd}} & \approx & \sum_{\bx \in \mathcal{M}_g} E_{\by^{j,s}|\mathcal{Y}^{s-1}} \left\{ p(\bx;\mathcal{Y}^{j,s})\left( 1-p(\bx;\mathcal{Y}^{j,s})\right) \right\} \Delta, \nonumber \\
    p(\bx;\mathcal{Y}^{j,s}) &=& P(\xi_a(\bx) \leq t_a, \xi_b(\bx) \leq t_b |\by^{j,s},\mathcal{Y}^{s-1}). \nonumber
\end{eqnarray}

Note that this strategy gives a sequential conditional version of the
formula in Eq. (\ref{sur}). Now $\mathcal{Y}^{s-1}$ is available, and
the expectation is with respect to the conditional density
$\pi(\by^{j,s}|\mathcal{Y}^{s-1})$. A similar closed form calculation
for expected IBV is hence applicable in Eq. (\ref{critSEQ}), using the
updated GP model from step $s-1$.  Once the data is collected for the
best design, the GP model is updated again. The mean $\bm_{s}$ and
covariance matrix $\bS_{s}$ are used to compute the next design at
stage $s+1$, and so on.

Even though this myopic strategy is non-anticipative, it still gives a
reasonable approach for creating designs in many
applications. Moreover, it is easily implemented on-board an AUV,
using the efficient approach for data updating of the GP model and the
calculation of the closed form expected IBV expressions for each
subsequent trajectory.

\subsection{Look-ahead trajectory planning}
\label{LA}

We now extend the myopic strategy to a look-ahead strategy which is
optimal when one can gather only two more rounds of measurements. In
addition to the next round of measurements $\by^{j,s}$, this
look-ahead strategy, anticipates the subsequent design $j_2$ with
data $\by^{j_2,s+1}$, when choosing the current design $d^{j,s}$.  The
selected design is:
\begin{eqnarray}\label{critLA}
    d^{*,s} &=& \mbox{argmin}_{j \in \{1,\ldots,J\}} \left\{ U_{la,\mbox{upd,j}} \right\},  \\
    U_{la,\mbox{upd},j} & = &  E_{\by^{j,s}|\mathcal{Y}^{s-1}} \left\{ \mbox{argmin}_{j_2 \in \{1,\ldots,{J}_2\}} \left[ V_{la,\mbox{upd},j_2} \right] \right\}, \nonumber \\
V_{la,\mbox{upd},j_2} & \approx & \sum_{\bx \in \mathcal{M}_g} E_{\by^{j_2,s+1}|\mathcal{Y}^{j,s}} \left\{ p(\bx;\mathcal{Y}^{j,s+1})\left( 1-p(\bx;\mathcal{Y}^{j,s+1})\right) \right\} \Delta, \nonumber \\
    p(\bx;\mathcal{Y}^{j,s+1}) &=& P(\xi_a(\bx) \leq t_a, \xi_b(\bx) \leq t_b |\by^{j_2,s+1},\mathcal{Y}^{j,s}). \nonumber
\end{eqnarray}
Here, $\mathcal{Y}^{j,s}=\{\by^{j,s},\mathcal{Y}^{s-1}\}$ and
$\mathcal{Y}^{j,s+1}=\{\by^{j_2,s+1},\mathcal{Y}^{j,s}\}$ represent
the sets of data variables, and the expectations are with respect to
the conditional densities $\pi(\by^{j,s}|\mathcal{Y}^{s-1})$ and
$\pi(\by^{j_2,s+1}|\by^{j,s},\mathcal{Y}^{s-1})$.

@@

Even though the strategy looks at two stages, it is only used to find
the current best design. When data is collected, the GP model is
updated, and the mean $\bm_{s}$ and covariance matrix $\bS_{s}$ are
used to compute the next design at stage $s+1$, now anticipating what
stage $s+2$ could offer, and so on.

This look-ahead approach is much more computationally demanding than
the myopic strategy, and for practical implementation we prune paths
in the evaluation of Eq. (\ref{critLA}). This means that we do not
compute all possible branches of the first two stages, as they are
indicated in Figure \ref{fig:PathSelOpt}. Instead, we use the myopic
strategy to rank the three best designs on the first stage alone, and
for each of these we undertake the look-ahead calculations.

\section{Simulation study}
\label{sec:simulations}

We now show results from tests performed to study the properties of
different static and sequential survey designs, in a realistic
simulated case. The context is mapping a freshwater plume defined by a
temperature and salinity gradient using an AUV.

@@

We use a bivariate GP model for temperature and salinity, where we
specify the prior mean:
\begin{equation}\label{m}
    \bmu(\bx)=E 
    \begin{bmatrix}
    \xi_a(\bx) \\
    \xi_b(\bx) 
    \end{bmatrix}=\begin{bmatrix} \mu_{a}(\bx)\\
\mu_{b}(\bx)
\end{bmatrix} 
= \begin{bmatrix} \beta_{a,0} + \beta_{a,1} x_{\mbox{East}} \\
\beta_{a,1} + \beta_{b,1} x_{\mbox{East}}
\end{bmatrix}.
\end{equation}
In this simulation setup, which is motivated by a real case, we expect
the eastern part of the domain to be cold freshwater. This situation
mimics that of a river mouth entering from the south, and the water
masses are pulled to the east. The actual trends are, in practice, set
from preliminary information or runs of oceanographic models \kc{a
  citation of such model use, e.g. the JFR paper might be in order
  here} solving differential equations for hydrodynamic flow. In this
example the mean is specified by $\beta_{a,1}=0.065$ and
$\beta_{b,1}=0.1$. Further, the intercepts \kc{Whats an intercept?} of
the temperature and salinity variables are set to $\beta_{a,0}=5.8$
and $\beta_{b,1}=29.0$.

@@

The covariance is assumed to be stationary and separable for the two variables. 
The $2 \times 2$ pointwise covariance matrix is set to:
\begin{equation}\label{v0}
\bSigma(\bx)=\mbox{Var} 
\begin{bmatrix}
    \xi_a(\bx) \\
    \xi_b(\bx) 
    \end{bmatrix}=
\begin{bmatrix}
0.25^2 & 0.6 \cdot 0.25^2 \\
0.6 \cdot 0.25^2 & 0.25^2
\end{bmatrix}.
\end{equation}
The spatial correlation is of a Matern type \kc{? Define or cite} giving:
\begin{equation}\label{v}
\mbox{Corr} 
\left(
\begin{bmatrix}
    \xi_a(\bx) \\
    \xi_b(\bx) 
    \end{bmatrix},
    \begin{bmatrix}
    \xi_a(\bx') \\
    \xi_b(\bx') 
    \end{bmatrix}
    \right)
    = \begin{bmatrix}
1 & 0.6  \\
0.6  & 1
\end{bmatrix}(1+\phi |\bx-\bx'|)\exp (-\phi |\bx-\bx'|),
\end{equation}
and $\phi=0.3$ indicates an effective correlation range of about $1200$ m. 
%Figure \ref{fig:stat_design} shows the contour lines for the EP for the reference. We notice the trend of increasing salinity and temperature to the west.  
We study sensitivity to the parameter specification in the results
below.  One realization of the random fields representing salinity and
temperature is shown in Figure \ref{fig:true_temp} and
\ref{fig:true_sal}. The true ES from this prior realization is shown
in Figure \ref{fig:ESet}.

\begin{figure}[!ht]
  \centering
  \subfigure[Simulated temperature.]{\includegraphics[height = 0.40\textwidth]{Figures/sim/true_temp.pdf}\label{fig:true_temp}}
  \hfill
  \subfigure[Simulated salinity.]{\includegraphics[height = 0.40\textwidth]{Figures/sim/true_sal.pdf}\label{fig:true_sal}}
  \hfill
  \subfigure[Excursion set given the realization of temperature and salinity.]{\includegraphics[height = 0.40\textwidth]{Figures/sim/es_ts_true.pdf}\label{fig:ESet}}
  \hfill
  \subfigure[Excursion probabilities given collected data using the
  `static\_north" strategy.]{\includegraphics[height = 0.40\textwidth]{Figures/sim/es_ts_posterior.pdf}\label{fig:Eprob}}
  \caption{\ref{fig:true_temp} and \ref{fig:true_sal} show
    realisations of the simulated temperature and salinity fields, as
    well as the associated excursion set \ref{fig:ESet}.
    \ref{fig:Eprob} shows the estimated excursion probabilities after
    performing data collection along the N-S survey line.} 
\label{fig:realisations}
\end{figure}

@@

We compare three different static designs denoting
\texttt{static\_north}, \texttt{static\_east} and
\texttt{static\_zigzag} with the three sequential approaches
\texttt{naive}, \texttt{myopic}, and \texttt{look-ahead}. The static
sampling paths are pre-determined and cannot be changed on the basis
of data, representing the archetypical pre-planned observation
strategies used in typical AUV operational survey designs.

For a fixed survey length, the expected variance after static
surveying is analytically available (equal coverage each time), but
here we focus on the properties from the sequential designs which are
evaluated using Monte Carlo sampling over several replicates of
realizations from the prior, and conducting simulated sequential
surveying for each one, making the result dependent on observed
data. For comparison, the same setup is used for static
designs. Figure \ref{fig:Eprob} shows the conditional EP, given data
gathered along the north-south survey line for the realization
indicated in these displays. In the Monte Carlo replicates, such
results are repeatedly computed to approximate the expected variance
reduction in the ES. Along with this criterion we also compare
predictive performance measured by root mean square error (RMSE) in
the temperature and salinity fields and variance reduction in these
two variables. It is important to note that the objective function
used by the agents is focused on reducing the expected variance in the
ES, but we nevertheless hope to achieve good predictive performance
for criteria such as RMSE as well. Another non-statistical criteria
that is important for practical purposes is the computational time of
the strategy, as this will impact the performance of the approach in
the field for an embedded system. In Fig. \ref{fig:wp_graph}, the
sequential sampling agents start at the center East-West coordinate at
the southern end of the domain (node 53).

\begin{figure}[t]
\centering
\includegraphics[width=0.50\textwidth]{Figures/sim/wp_graph_paper.pdf}
\caption{The equilateral waypoint graph used to discretize the
  trajectory choices over the $31\times31$ grid.}
\label{fig:wp_graph}
\end{figure}

Maintaining the example above, the AUV will move along edges in the
Fig. \ref{fig:wp_graph} waypoint graph while the vehicle gathers
measurements. These are assimilated in the GP model, before an
evaluation of the next node to sample is conducted at the end of the
edge. The paths will differ for the various sequential designs due to
the individual design criteria and the simulated variability in the
environment among the replicates.

@@

\section{Case study - Mapping a river plume}
\label{sec:case_study}

To demonstrate the applicability of multivariate EP to informative
oceanographic sampling, we present a case study mapping river plumes
with an AUV.  This experiment was performed in Trondheim, Norway,
where the Nidelva river (Figure \ref{fig:nidelven}) meets water from
the Trondheimsfjord. The experiment is focused along the eastern
shoreline due to the aforementioned Coriolis
force.%, see Fig. \ref{fig:map}.
The frontal zone therefore runs parallel to the eastern shore as
indicated in purple as it enters the fjord. The experiments were
conducted in late Spring 2019, and because of snow melt with
consequent fresh river water substantially colder than fjord waters.

\subsection{Model specification}
